Choose a License

### 2024-12-04 choose_license.qmd
Incorporate feedback from Felix and Malika

---
title: "Choose a License"
engine: knitr
---

@@

Researchers are only allowed to share material they have the necessary rights to. This includes copyrights, which are granted by free/open copyright licenses. These licenses often require to attribute the authors, indicate whether changes were made, and provide the text of the license, among other things. In addition, researchers should share any work by their own under such a license.

@@

A license is a legal document that regulates what _others_ are allowed to do with a copyright-protected work -- the licenses we discuss are non-exclusive licenses and do not limit the author or copyright holder in their rights. And while you _could_ write your own license, there are already many pre-formulated licenses available to choose from and to apply to your work.^[In fact, you should not create your own license if the purpose is to share your work freely and openly with others. By using a boilerplate license others know what to expect, also because they have been [tested](https://sfconservancy.org/copyleft-compliance/) in [courts](https://legaldb.creativecommons.org/).]

@@

::: {#wrn-exclusive-rights .callout-warning}
### Giving Publishers Exclusive Rights

Some journals offer to publish your article under a Creative Commons license, but still demand an exclusive publishing and distribution license or a copyright assignment from you. This would give them more rights than the readers of the article have through the respective Creative Commons license [@Rumsey2022] and exceeds by far what is necessary to make publication possible [@Suber2022]. Consequently, authors should oppose this practice and grant publishers the same rights that every other reader of the article has. If your chosen publisher insists on an exclusive license, you may at least retain the copyright for your figures -- follow the guide "Retaining copyright for figures in academic publications to allow easy citation and reuse" by @Elson2016 to learn how to do that.

@@

First, if you adapt (i.e., modify, remix, build on) a work by others you need to determine if it is provided to you under a free/open license.[You can determine whether a license is free/open by searching for its name in the [SPDX License List](https://spdx.org/licenses/) and looking for at least one `Y` in the two columns _FSF Free/Libre?_ and _OSI Approved?_]{.aside} If yes, we recommend you to make your contribution available under the same license.^[Copyleft licenses even require you to choose the same or a compatible license.] For example, if you adapt code published in another paper, choose the same license for your modifications. The same applies if there are strong community norms to use a particular free/open license^[Of course, this is only a heuristic and there might be good reasons to deviate from community norms.] or if a particular license choice is mandated by your funder. Importantly, as discussed before, you are generally not allowed to adapt a work _not_ published under a free/open license.

@@

::: {#fig-flowchart-simple}
```{mermaid}
flowchart TB
  start("We want<br>to choose<br>a license.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>concerning license?")

@@

One should take special care when applying license to the following types of works:

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and TeX packages:__ If a template or, for this purpose, a $\TeX$ package, is licensed under a copyleft license such as the (A)GPLv3, every work that is a derivative of the template or that uses the $\TeX$ package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the (A)GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to be licensed under [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or another license).
:::

@@

In most cases, the output of software, like images or tables, does not depend on the software's license.^[The exception is when the software is part of the output [see @apsillers2019].] Therefore, if you use an R package under a copyleft license to create a figure, you are likely the copyright owner. However, if the output is based on data, it can be considered a derivative work of the data and the license of the data also applies. For example, maps may be considered as a derivative of the geographic data they are based on.

@@

- __Attribution__ means whether recipients of your work are required to provide attribution to you. For software licenses, the attribution is part of a copyright notice. Note that even if attribution is not a requirement of the license, good scientific practice demands that appropriate citations are made.
- __State changes__ means that recipients need to indicate if changes were made.
- __Anti-DRM__ means that when others share your work they are not allowed to apply technological measures restricting anything that the license permits. DRM is an abbreviation for **d**igital **r**ights **m**anagement.

@@

::: {#fig-flowchart-non-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for software,<br>writing, image, audio,<br>or video.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>concerning license?")

@@

::: {#fig-flowchart-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for data.") --"We adapted a database by<br>others shared under a<br>free/open license."--> use_existing_license_db["<em>Use its license(s)<br>for content and database</em>"]
  start --"We created a database<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>concerning license?")

>>>>

---
title: "Choose a License"
lang: en
format:
  html:
    mermaid:
      theme: neutral
engine: knitr
---

@@

Researchers are only allowed to share material they have the necessary rights to, for example, through a free/open license. These licenses often require to attribute the authors, indicate whether changes were made, and provide the text of the license, among other things. In addition, researchers should share any work by their own under such a license.

@@

::: {#nte-situation-lmu .callout-note collapse="true"}
### For researchers at LMU

If you are an employee at LMU and your funder does not mandate the use of a particular license,
the decision of which license to use lies with the holder of your chair (_Lehrstuhlinhaber\*in_).
In the (rare) case that the software you developed has a technical application,
that it falls under the legal definition of an invention,
you are required to officially report your invention to the university which decides on its release.
However, the [IP strategy](https://www.serviceportal.verwaltung.uni-muenchen.de/services/forschungsservices/_downloads/07_ip_strategy.pdf) of the LMU highlights that
"LMU can make the greatest contribution to the value chain by focusing on the production of outstanding results in basic research and by helping to **make the results publicly accessible**.
It follows that the commercial exploitation of research results has an accessory role".
Furthermore, "as a rule, the inventor is entitled to the discretionary use of invention- and research-based results".
Taken together, LMU emphasizes the importance of openness of research results and the freedom of researchers to share their findings widely.

Only if the university claims the invention after you reported it, you may not apply a copyright license that contains a patent grant. In all other cases you are free to license your research project as you like.

More information about the invention disclosure process can be found in the [service portal](https://www.serviceportal.verwaltung.uni-muenchen.de/internationals/research_transfer/intellectual_property/index.html).
More information about software patents are provided by the FSFE: [Software Patents in Europe](https://fsfe.org/activities/swpat/swpat.en.html).
:::

A license is a legal document that regulates what _others_ are allowed to do with a copyright-protected work -- the licenses we discuss do not limit the author or copyright holder in their rights. And while you _could_ write your own, there are already many pre-formulated licenses available to choose from and to apply to your work.^[In fact, you should not create your own license if the purpose is to share your work freely and openly with others. By using a boilerplate license others know what to expect, also because they have been [tested](https://sfconservancy.org/copyleft-compliance/) in [courts](https://legaldb.creativecommons.org/).]

@@

::: {#cau-exclusive-rights .callout-caution}
### Giving Publishers Exclusive Rights

When you have your article published under a Creative Commons license, some publishers still demand an exclusive publishing and distribution license or a copyright assignment. This would give them more rights than the readers of the article have through the chosen Creative Commons license [@Rumsey2022] and exceeds by far what is necessary to make publication possible [@Suber2022]. Consequently, authors should oppose this practice and grant publishers the same rights that every other reader of the article has. If your chosen publisher insists on an exclusive license, you may at least retain the copyright for your figures -- follow the guide "Retaining copyright for figures in academic publications to allow easy citation and reuse" by @Elson2016 to learn how to do that.

@@

First, if you adapt (i.e., modify, build on) a work by others you need to determine if it is provided to you under a free/open license.[You can determine whether a license is free/open by searching for its name in the [SPDX License List](https://spdx.org/licenses/) and looking for at least one `Y` in the two columns _FSF Free/Libre?_ and _OSI Approved?_]{.aside} If yes, we recommend you to make your contribution available under the same license.^[Copyleft licenses even require you to choose the same or a compatible license.] For example, if you adapt code published in another paper, choose the same license for your modifications. The same applies if there are strong community norms to use a particular free/open license^[Of course, this is only a heuristic and there might be good reasons to deviate from community norms.] or if a particular license choice is mandated by your funder. Importantly, as discussed before, you are generally not allowed to adapt a work _not_ published under a free/open license.

@@

::: {#fig-flowchart-simple}
```{mermaid}
flowchart TB
  start("We want<br>to choose<br>a license.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")

@@

One should be cautious about the restrictions of licenses applied to the following types of works:

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and LaTeX packages:__ If a template or, for this purpose, a LaTeX package, is licensed under a copyleft software license such as the [AGPLv3](https://choosealicense.com/licenses/agpl-3.0/), every work that is a derivative of the template or that uses the LaTeX package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to licensed under CC\ BY-SA\ 4.0 (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or even another license).
:::

@@

In most cases, the output of software, like images or tables, does not depend on the software's license. Therefore, if you use an R package under a copyleft license to create a figure, you are likely the copyright owner. However, if the output is based on data, it can be considered a derivative work and the license of the data also applies. For example, maps may be considered as a derivative of the geographic data they are based on.

@@

- __Attribution__ means whether recipients of your work are required to provide attribution to you. For software licenses, this is called a copyright notice. Note that even if attribution is not a requirement of the license, good scientific practice demands that appropriate citations are made.
- __State changes__ means that recipients need to indicate if changes were made.
- __Anti-DRM__ means that when others share your work they are not allowed to apply technological measures restricting anything that the license permits. DRM is an abbreviation for **d**igital **r**ights **m**anagement.

@@

::: {#fig-flowchart-non-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for software,<br>writing, image, audio,<br>or video.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")

@@

::: {#fig-flowchart-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for data.") --"We adapted a database by<br>others shared under a<br>free/open license."--> use_existing_license_db["<em>Use its license(s)<br>for content and database</em>"]
  start --"We created a database<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")
